Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-11-30 manuscript.Rmd
update 2022-11-30 15:30:14

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LCA"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis is a family of techniques for identifying groups in data based
  on a parametric model. Its examples include mixture models, latent class
  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on
  the `OpenMx` R-package.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Despite its popularity, there is a lack of standards for estimating and
reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
reporting guidelines for a specific type of LCA known as latent growth
models, general reporting guidelines for all LCA are still lacking. This
complicates manuscript review and assessment of the quality of published
studies, and introduces a risk of misapplications of the technique. The
present paper seeks to address this gap in the literature by suggesting
updated guidelines for estimation and reporting of LCA, based on current
best practices. Importantly, in order to lower the barrier of entry and
ensure reproducibility of all examples, this paper exclusively relies on
free, open source software in R. Our goal is to make best-practices in
LCA widely accessible.

@@

Latent class analysis is a group of methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership. The concept of LCA can be understood in different ways. A
mixture model assumes that the study population is composed of $K$
subpopulations or classes. It further assumes that the observed data are
a mixture of data generated by class-specific models. The simplest
univariate model is a normal distribution, which can be described with
two parameters: the mean and the variance. Commonly, the same model is
estimated across all classes, but with different parameters for each
class (i.e., class-specific means and variances). Mixture modeling then
estimates both the parameters for each class, and the probability that
an individual belongs to each class. This way, describing the whole
sample with a single mean and variance is a special case where we assume
the sample is homogeneous in function of these parameters.

@@

LCA is like confirmatory factor analysis, except that the continuous
latent variable is substituted with a categorical latent variable. Both
assume that the indicators are related, and these relations can be
represented by the underlying latent variable. One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes. While factor analysis seeks to describe the
relations between variables, LCA seeks to describe each unobserved group
and the differences between them. In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm. In line with this
perspective, LCA is sometimes described as "model-based clustering"
[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model. Specifically, the
posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data
under given the class-specific model. For these differences, 
LCA is a clustering method, that corrects for (random) measurement error, 
while clustering algorithms do not correct for. This is something LCA has 
in common with factor analysis, as both model an underlying latent variable, 
which accounts for the communalities between indicators 
and correcting for measurement error.

Finally, in the context of machine learning, LCA can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable --
true class membership -- is not known, and the term *classification*
refers to the fact that the algorithm is predicting a categorical
outcome -- class membership.

@@

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations
are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2009]. As an example,
a LPA of ocean plastic particles found two classes of particles based on length
and width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape [@alkema_maximizing_2022]. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.

@@

The term latent class analysis originally referred to cases where the
observed indicators were categorical [@collins_latent_2009]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*. Latent class models
with ordinal indicators are parameterized differently from mixture
models. One common parameterization assumes that each categorical
variable reflects an underlying standard normal distribution. The
parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.

This paper will primarily focus on mixture models and special cases
thereof, although most of the suggested guidelines are applicable to all
LCA methods.

@@

There are several use cases for which LCA methods are suitable. One
such application is to test a theory that postulates the existence of a
categorical latent variable. For instance, *identity status theory*
posits that, at any given point in time, adolescents reside in one of
four identity statuses [@marcia_development_1966; @marcia_identity_1980]. 
LCA can be used to test whether these four statuses
can be recovered based on observed indicators (e.g., self-reported
identity exploration and commitment). If the results indicate that the data
are better described by a different number of classes, or that the
four-class solution does not correspond to the predicted pattern of
responses on the indicators, then the theory may be called into
question. A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.
A LCA revealed three identity profiles: full integration (high belonging to 
heritage, national and regional identity), national integration (high heritage
and national identity with low regional identity), and weak separation
(high heritage identity with neutral attitudes towards the nation and region).

>>>>

---
title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "3"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
  - name          : "Daniel Anadria"
    affiliation   : "1"
    corresponding : no
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "d.anadria@uu.nl"
affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"
  - id            : "3"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Its examples include mixture models, latent class analysis with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
wordcount         : "9001"
bibliography      : "mixture.bib"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

@@

Despite its popularity, there is a lack of standards for estimating and
reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
reporting guidelines for a specific type of LCA known as latent growth
models, general reporting guidelines for best practices in LCA are still
lacking.
This introduces a risk of misapplications of the technique,
and complicates manuscript review and assessment of the quality of published studies.
The present paper seeks to address this gap in the literature by introducing updated guidelines for estimation and reporting on latent class analysis,
based on current best practice. 
Importantly, in order to lower the barrier of entry to latent class analysis
and ensure reproducibility of all examples,
this paper exclusively uses free, open source software for latent class analysis in R.
To this end, new functionality was developed in the R-package `tidySEM` [REF tidysem].

@@

Latent class analysis is a group of methods for estimating unobserved
groups based on a parametric model of observed indicators of group
membership. The concept of LCA can be understood in different ways. A
mixture model assumes that the study population is composed of $K$
subpopulations or classes.
It further assumes that the observed data are
a mixture of data originating from those subpopulations;
hence the name mixture model.
Consider the simplest possible
univariate "model", which is a normal distribution.
This model has
two parameters: the mean and the variance.
Latent class analysis aims to estimate the values of those parameters across a known number of classes $k$,
as well as the probability of belonging to classes $1 \ldots k$ for every individual $i$.
Commonly, the same model is estimated across all classes,
but with different parameters for each class (i.e., class-specific means and variances).
From this perspective, reporting the mean and variance of a variable is a special case of a mixture model with $k = 1$ class.

@@

Another way to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis,
except that the continuous latent variable is substituted with a categorical latent variable.
One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes. While factor analysis seeks to describe the
relations between variables, LCA seeks to describe each unobserved group
and the differences between them. In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

When the focus is on each individual's estimated class membership,
LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.
These posterior class probabilities can be used to weight follow-up analyses,
or individuals can be assigned to classes based on their highest class probability.
If the classes are clearly distinct, measurement error is low, and the latter approach is probably acceptable.
This perspective on mixture modeling is sometimes referred to as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016]. 
Many clustering algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model.
Since a parametric model estimates only a fixed number of parameters,
it can be a relatively parsimonious solution compared to non-parametric techniques.

Finally, in the context of machine learning, LCA can be considered as an
*unsupervised classification* method [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable --
true class membership -- is not known, and the term *classification*
refers to the fact that the algorithm is predicting a categorical
outcome -- class membership.

@@

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for,
the residual covariances/correlations between indicators are
assumed to be zero.
In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2009].
Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
A class of larger particles with a low correlation between length and width,
and a class of smaller particles with a high correlation between length and width.
The reason for this difference in correlations makes theoretical sense:
the large particles were heterogenous in shape,
whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements [@alkema_maximizing_2022].

@@

The term latent class analysis originally referred to cases where the
observed indicators were ordinal [@collins_latent_2009]. 
Nowadays, it is more commonly used
as an umbrella term. To prevent ambiguity, the special case where
indicators are of binary or ordinal measurement level might be described
as *latent class analysis with ordinal indicators*.
Latent class models with ordinal indicators are parametrized differently from mixture models.
One common parametrization assumes that each categorical variable reflects an underlying standard normal distribution.
The parameters are "thresholds" that correspond to quantiles of a standard
normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
are estimated based on the proportion of individuals in each of the
response categories of the indicator variable. For example, a binary
indicator has a single threshold that distinguishes the two response
categories. If responses are distributed 50/50, then the corresponding
threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
then the resulting threshold would be $t_1 = 0.25$.
Although this paper primarily focuses on latent class analysis with continuous indicators,
most of the suggested guidelines are equally applicable to ordinal indicators.

@@

There are several use cases for which LCA methods are suitable. 
Although latent class analysis is often discussed as an exploratory analysis technique,
it can also be used in a confirmatory manner.
Given that latent class analysis is similar to confirmatory factor analysis,
but with a categorical latent variable, it can be used to similar ends:
To assess whether a theoretical measurement model holds.
This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
One challenge is that, unlike confirmatory factor analysis,
LCA does not provide absolute fit indices that can be used to perform inference.
Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
either qualitatively or quantitatively, using significance tests for pre-specified values.
A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.
